---
title: "LIFE748 - ML and data analysis"
author: "Kirsten Leigh Nutter"
format: html
editor: visual
---

```{# AI Declaration statement: I used GenAI tools in the preparation of this work}
  
# Claude v4.6	
# I used Claude for part 2 analysis only, to aid with the code for steps 6, 7 and 8 of the R .qmd file. Specifically, I used it for ROC curve to compare svm kernel performance, pls-da analysis (not included in final report), and cross validation to implement a k-fold cross validation approach to compare across random splits. I have rewritten the code that was provided in responses out myself as not all of it was necessary, I have added parts/taken parts of code out to align with the aims of the specific steps. Please see prompt/response file for more information on use for each step.

```

## Introduction

In this tutorial, we will analyze a simulated RNA-seq dataset. Our goal is to identify Differentially Expressed Genes (DEGs) between two conditions (Condition A and Condition B), visualize the results, and use both unsupervised and supervised machine learning techniques to classify the samples based on their gene expression profiles.

## Step 0: Load the required libraries

```{r warning=FALSE, message=FALSE }

# Install BiocManager

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.22")

BiocManager::install(c("S4Vectors", "IRanges", "SummarizedExperiment", "GenomicRanges"))

# Install BiocManager packages
BiocManager::install(c("DESeq2", "EnhancedVolcano", "pheatmap"), ask = FALSE, update = TRUE)

# Install CRAN packages
install.packages(c("ggplot2", "caret", "MASS", "e1071"))

# Load libraries 
library(DESeq2) # Bioconductor
library(ggplot2) # CRAN
library(EnhancedVolcano) # Bioconductor
library(pheatmap) # Bioconductor
library(caret) # CRAN
library(MASS) # CRAN
library(e1071) # CRAN

```

If these packages are not already available, you will need to install them either from CRAN (install.packages('package_name')), or from Bioconductor (BiocManager::install('package_name')

## Step 1: Load the Count Matrix

First, we load the raw count matrix and set up the sample metadata (phenotype data). We assume the first 25 columns belong to Condition A, and the next 25 belong to Condition B.

```{r}
counts <- read.csv("dataset_LIFE748_3.csv", row.names = 1)

colData <- data.frame(
  row.names = colnames(counts),
  Condition = c(rep('CondA',25),rep('CondB',25))
)

```

## Step 2: Normalize and Identify DEGs with DESeq2

We will use `DESeq2` to normalize the library sizes and perform statistical testing to identify which genes are differentially expressed between Condition B and Condition A.

```{r}
# Create DESeq2 dataset object
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = colData,
                              design = ~ Condition)

# Run DESeq2 pipeline
dds <- DESeq(dds)

# Extract the results (Condition B vs Condition A)
res <- results(dds)

# Convert object into a dataframe
res <- as.data.frame(res)

# Extract normalised counts for downstream machine learning
vsd <- vst(dds, blind = FALSE) # Variance stabilising transformation
normalized_counts <- assay(vsd)

```

## Step 3: Report DEGs (Table and Volcano Plot)

We define statistically significant DEGs as those with an adjusted p-value (FDR) \< 0.05 and an absolute log2 Fold Change \> 1.

```{r}

# Filtering for significant DEGs, those with FDR < 0.05 and absolute log2FC > 1
sigDEGs <- res[which(res$padj < 0.05 & abs(res$log2FoldChange) > 1), ]

# Order by adjusted p-value
sigDEGs <- sigDEGs[order(sigDEGs$padj), ]

# Summary counts
cat("Total DEGs:", nrow(sigDEGs), "\n")
cat("Upregulated (log2FC > 1):", sum(sigDEGs$log2FoldChange > 1), "\n")
cat("Downregulated (log2FC < -1):", sum(sigDEGs$log2FoldChange < -1), "\n")

# Top 20 DEGs
head(sigDEGs[, c("baseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj")], 20)

# baseMean = How highly expressed the gene is
# log2FoldChange = How much it changes between conditions
# lfcSE = How confident DESeq2 is in that change, standard error
# stat = FC divided by its uncertainty (signal/noise)

write.csv(sigDEGs, "DEGs_significant.csv")

```

Generate a volcano plot to visualise the results

```{r fig.width=12, fig.height=11}

EnhancedVolcano(res,
  lab        = rownames(res),                 # gene labels
  x          = 'log2FoldChange',              # x-axis
  y          = 'padj',                        # y-axis (FDR-adjusted p-value)
  title      = 'Condition A vs Condition B',
  subtitle   = 'FDR < 0.05, log2FC > 1',
  pCutoff    = 0.05,                          # horizontal threshold line
  FCcutoff   = 1.0,                           # vertical threshold lines
  pointSize  = 3,
  labSize    = 3.5,
  col        = c('red',                       # Not significant
                 'palevioletred',             # FC only
                 'orange',                    # p-value only
                 'green'),                    # significant both fc and p val
  colAlpha   = 0.6,
  legendPosition = 'right',
  widthConnectors = 0.4,
  max.overlaps = 100                    
)

```

## Step 4: Isolate Differentially Expressed Genes

To reduce noise and dimensionality for our machine learning algorithms, we will subset our normalized count matrix to only include the statistically significant DEGs identified above.

In order to make sure we don't lose too much information, let's relax a bit the thresholds for significance

```{r}

# Relaxed thresholds: FDR < 0.1 and |log2FC| > 0.5
degs_relaxed <- res[which(res$padj < 0.1 & abs(res$log2FoldChange) > 0.5), ]

cat("DEGs with strict thresholds (FDR<0.05, |FC|>1):", 
    nrow(res[which(res$padj < 0.05 & abs(res$log2FoldChange) > 1), ]), "\n")
cat("DEGs with relaxed thresholds (FDR<0.1, |FC|>0.5):", 
    nrow(degs_relaxed), "\n")


# Subset gene names from relaxed DEG list
deg_genes <- rownames(degs_relaxed)

# Subset the VST-normalised matrix to only the DEG features
ml_matrix <- normalized_counts[deg_genes, ]

dim(ml_matrix)

ml_matrix[1:5, 1:5]



# Transpose to make rows = samples and columns = genes
ml_data <- as.data.frame(t(ml_matrix))

deg_counts_scaled <- scale(ml_data)

# Add condition labels as the target variable
ml_data$Condition <- colData$Condition

dim(ml_data)        
table(ml_data$Condition)  

# Condition to a factor
ml_data$Condition <- as.factor(ml_data$Condition)

```

## Step 5: Unsupervised Clustering

Let's see if the samples naturally group by their condition without explicitly telling the algorithm which sample belongs to which group. We will use Hierarchical Clustering and K-means.

## k-means

```{r}

df <- ml_data[, !colnames(ml_data) %in% "Condition"]
df_scaled <- as.data.frame(scale(df))


set.seed(123)
out <- kmeans(df_scaled, centers = 2, iter.max = 15, nstart=5) 
out

# % needs to be above 60% (change center number until this has been reached) 

df_scaled$clusters <- as.factor(out$cluster)
df_scaled$Condition <- ml_data$Condition

pca <- prcomp(df_scaled[, !colnames(df_scaled) %in% c("clusters","Condition")])
plot_df <- data.frame(
  PC1       = pca$x[,1],
  PC2       = pca$x[,2],
  Cluster   = df_scaled$clusters,
  Condition = df_scaled$Condition
)


ggplot(plot_df, aes(x=PC1, y=PC2))+
  geom_point(aes(col=Cluster))+
  theme(panel.background = element_rect(fill = "white",colour = "black", 
                                        linewidth = 1,linetype = "solid"),
        panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
        text = element_text(size=20,colour = "black",face = "bold"),
        legend.key=element_blank())+
  ggtitle("K-means clusters vs true conditions")


# Elbow test

gene_cols <- !colnames(df_scaled) %in% c("clusters", "Condition")

k.max <- 10
wss <- sapply(1:k.max, 
              function(k){kmeans(df_scaled[, gene_cols], k, nstart=5, iter.max = 15 )$tot.withinss})
plot(1:k.max, wss,
     type="b", pch = 19, frame = FALSE, 
     xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")
abline(v = 2, lty = 2, col = "palevioletred")

# further clustering is needed to show how similar each data point is to the others - hierarchical clustering for a dendrogram, this doesnt give you clusters but it is more adpatable than k-means by giving the relationship between the points (distance matrix)
```

## Hierarchical Clustering

```{r}
dist_mat <- dist(df_scaled[, gene_cols], method = "euclidean")


hc_average<- hclust(dist_mat, method = 'average')
plot(hc_average)
#rect.hclust(hc_average, h = 1.3, border = 2:6)
rect.hclust(hc_average, k = 2, border = 2:6)  
abline(h = 1.3, col = 'palevioletred')

```

Visualise the results (HINT: an heatmap would be perfect for this).

```{r}
set.seed(123)

km <- kmeans(deg_counts_scaled, centers = 2, nstart = 25)

km$cluster

table(Cluster = km$cluster, Condition = colData$Condition)
```

```{r}

annotation_col <- data.frame(Cluster = as.factor(km$cluster), Condition = colData$Condition)
rownames(annotation_col) <- rownames(colData)                             

# Heatmap of DEGs (all samples)
library(pheatmap)
pheatmap(t(scale(t(ml_matrix))),
         annotation_col = annotation_col, 
         show_rownames = FALSE,
         main = "Heatmap of DEGs")

```

## Step 6: Supervised Learning (LDA and SVM)

Next, we train classifiers to predict the condition of a sample based on its DEG expression profile. We will split the data into a training set (70%) and a testing set (30%).

```{r}
ml_data <- as.data.frame(t(ml_matrix))
ml_data$Condition <- as.factor(colData$Condition)
dim(ml_data)
table(ml_data$Condition)

set.seed(122)
trainIndex <- createDataPartition(ml_data$Condition,p=0.7, list=FALSE,times=1)
df_train <- ml_data[trainIndex,] 
df_validate <- ml_data[-trainIndex,]
```

Train LDA (Linear Discriminant Analysis)

```{r}
model_LDA <- lda(Condition ~., data = df_train)
model_LDA

plot(model_LDA)

#p = [x1, y1]

# each point equi-distant from the centroid 
# becomes ellipsis shape of clusters (x and y)
# line between the two is the LDA line
# to discover linear combinations of given features 
# maximises seperation between classes and minimises variance within classes


pred <- predict(model_LDA, newdata=df_validate)
CM = confusionMatrix(data=as.factor(pred$class), as.factor(df_validate$Condition))
CM
```

Train SVM (Support Vector Machine)

```{r}
df_train$Condition <- as.factor(df_train$Condition)
df_validate$Condition <- factor(df_validate$Condition, levels = levels(df_train$Condition))

model_SVM <- svm(Condition ~ ., data=df_train)
summary(model_SVM)


pred <- predict(model_SVM, newdata=df_validate)

CM <- confusionMatrix(data = pred, reference = df_validate$Condition)
CM

#CM = confusionMatrix(data=as.factor(pred), as.factor(df_validate$Condition))
#CM

# svm is used if an lda line cannot be plotted e.g. if x points are surround by y in a circle (no clear seperation)

# svm enables us to plot the points using a different kernel (shape) to visualise the variance when plotted on a 2d graph, points are plotted on each side of optimal hyperplane.

# Kernels calculate similarity of pairs of data points

# Work out which kernel to use :

#LINEAR 
df_train$Condition <- as.factor(df_train$Condition)
model_SVM <- svm(Condition ~ ., data=df_train,kernel='linear')

pred <- predict(model_SVM, newdata=df_validate,type="class")
CM = confusionMatrix(data=as.factor(pred), as.factor(df_validate$Condition))
CM

#POLYNOMIAL
df_train$Condition <- as.factor(df_train$Condition)
model_SVM <- svm(Condition ~ ., data=df_train,kernel='polynomial')

pred <- predict(model_SVM, newdata=df_validate,type="class")
CM = confusionMatrix(data=as.factor(pred), as.factor(df_validate$Condition))
CM

#SIGMOID
df_train$Condition <- as.factor(df_train$Condition)
model_SVM <- svm(Condition ~ ., data=df_train,kernel='sigmoid')

pred <- predict(model_SVM, newdata=df_validate,type="class")
CM = confusionMatrix(data=as.factor(pred), as.factor(df_validate$Condition))
CM


# Acc > NIR P value should be < 0.05
# Mcnemars P value should be > 0.05

# linear P-Value [Acc > NIR] : 0.00647 - yes 
# linear Mcnemar's Test P-Value : 0.47950 - yes

# Polynomial P-Value [Acc > NIR] : 0.21198 - no, is not less than 0.05        
# Polynomial Mcnemar's Test P-Value : 0.07364 - yes

# Sigmoid P-Value [Acc > NIR] : 0.0009155 - yes
# Sigmoid Mcnemar's Test P-Value : 1.0000000 - yes


# from these results alone, polynomial can be removed as it is not better for predictions/similarity of pairs of data points than the baseline model

# between linear and sigmoid 
# Sigmoid results are stronger 

```

```{r}
# ROC curve plot to compare linear and sigmoid kernels for SVM 

library(e1071)
library(pROC)
library(ggplot2)

# Train both models with probability = TRUE
model_linear  <- svm(Condition ~ ., data = df_train, kernel = "linear",  probability = TRUE)
model_sigmoid <- svm(Condition ~ ., data = df_train, kernel = "sigmoid", probability = TRUE)

# Get probabilities
pred_linear  <- predict(model_linear,  newdata = df_validate, probability = TRUE)
pred_sigmoid <- predict(model_sigmoid, newdata = df_validate, probability = TRUE)

prob_linear  <- attr(pred_linear,  "probabilities")[, 2]
prob_sigmoid <- attr(pred_sigmoid, "probabilities")[, 2]

# ROC objects
roc_linear  <- roc(df_validate$Condition, prob_linear)
roc_sigmoid <- roc(df_validate$Condition, prob_sigmoid)

# Plot
ggroc(list(Linear = roc_linear, Sigmoid = roc_sigmoid), size = 1) +
  labs(title = "ROC Curve: Linear vs Sigmoid SVM",
       color = "Kernel") +
  annotate("text", x = 0.5, y = 0.4,
           label = paste("Linear AUC =",  round(auc(roc_linear),  3)), color = "palevioletred") +
  annotate("text", x = 0.5, y = 0.3,
           label = paste("Sigmoid AUC =", round(auc(roc_sigmoid), 3)), color = "steelblue") +
  theme_minimal()

# sigmoid has higher accuracy and kappa 
# an auc score of 1 for sigmoid indicates that by using this kernel in the sv model offers better discrimination between condition a and condition b, and whenever it identfies one condition, it is 100% confident it is not the other 
```

## Step 7: Compare Classifier Performance

Finally, we use confusion matrices to evaluate and compare how well LDA and SVM performed on the unseen test dataset.

```{r}
library(pls)
library(caret)
library(mixOmics)

df_train$Condition <- as.factor(df_train$Condition)

X <- apply(df_train[, -6], 2, as.numeric)
X <- as.matrix(X)
Y <- df_train$Condition

model_PLSDA <- plsda(X, Y, ncomp=2) #cv=10)
model_PLSDA

scores <- model_PLSDA$variates$X

plotdata <- data.frame(
  Comp1 = scores[,1],
  Comp2 = scores[,2],
  class = Y
)

ggplot(plotdata, aes(Comp1, Comp2, col = class)) +
  geom_point(size = 2) +
  theme_minimal()+
  scale_color_manual(values = c("steelblue", "palevioletred")) +
  labs(title = "PLS-DA Score Plot", x = "Component 1", y = "Component 2")

X_validate <- apply(df_validate[, -6], 2, as.numeric)
X_validate <- as.matrix(X_validate)

pred_PLSDA <- predict(model_PLSDA, X_validate, ncomp = 2)
pred_class <- pred_class <- pred_PLSDA$class$max.dist[, 2]
pred_factor <- factor(pred_class, labels = levels(df_train$Condition))

cm_PLSDA <- confusionMatrix(pred_factor, df_validate$Condition)
cm_PLSDA
```

## Step 8: Is this enough?

Is this enough to compare the classifiers? What if you use a k-fold cross-validation approach? Will you get similar results with different random splits?

This is not enough on its own to compare the classifiers as we have only tested on a single test and train. If a k-fold cross-validation approach was used, for example 5-fold, this would train on 4 folds and test on 1 to estimate how well each classifier performs on the unseen data to provide a better comparison. This would determine if performance results are stable across random splits.

```{r}
library(caret)
library(dplyr)

# defining a function for performance metrics for the models
prediction_metrics <- function(pred, truth, model_name, positive = "CondB") {
  cm1 <- confusionMatrix(pred, truth, positive = positive)# positive Condition B
  data.frame(Model=model_name, # for each mdel, creates a dataframe with all performance metrics:
             Accuracy = unname(cm1$overall["Accuracy"]), # overall accuracy measure
             Kappa = unname(cm1$overall["Kappa"]),# verall Kappa 
             Sensitivity = unname(cm1$byClass["Sensitivity"]), # metrics for class level 
             Specificity = unname(cm1$byClass["Specificity"]),
             Balanced_accuracy = unname(cm1$byClass["Balanced Accuracy"])
  )
}

set.seed(140)

folds <- createMultiFolds(ml_data$Condition, k = 5, times = 10) # 5 folds for each cross validation , which is then repeated 10 times 

crossvalidation <- list() # list to store metrcis per fold in each of the models
gene_log <- list() # list to store top selected genes for each of the folds

for (i in seq_along(folds)) { # for loop to loop through each fold 
  fold_train_index <- folds[[i]]
  fold_train <- ml_data[fold_train_index, , drop = FALSE] # get the train set for each fold in turn
  fold_test <- ml_data[-fold_train_index, , drop = FALSE] # get the test set for each fold in turn 
  fold_x <- fold_train[, setdiff(colnames(fold_train), "Condition"), drop = FALSE]
  fold_var <- sort(apply(fold_x, 2, var), decreasing = TRUE) # variance calculations using all columns except condition, for each feature in the training set
  fold_var <- fold_var[fold_var > 0] # only look at features with variance
  foldnfeatures <- min(10, length(fold_var)) # extracting top 10
  fold_features <- intersect(names(fold_var)[1:foldnfeatures], colnames(fold_test)) # and ensure that these are present in the test set

  gene_log[[i]] <- data.frame( # nsure that the features selected for each fold are recorded 
  Resample = names(folds)[i],
  Gene = fold_features
  )
  
  fold_train_x <- scale(fold_train[, fold_features, drop = FALSE]) # scaling both test and train features in the same way, which will stop leakage
  fold_centre <- attr(fold_train_x, "scaled:center")
  fold_scale <- attr(fold_train_x, "scaled:scale")
  fold_test_x <- scale(fold_test[, fold_features, drop = FALSE], 
                       center = fold_centre,
                       scale = fold_scale)
  
  fold_train_df <- data.frame(as.data.frame(fold_train_x), Condition = fold_train$Condition) # new dataframe that combines the scaled features into train and tst for modelling 
  fold_test_df <- data.frame(as.data.frame(fold_test_x), Condition = fold_test$Condition)
  
  # LDA 
  lda_cv_fit <- lda(Condition ~., data = fold_train_df)
  lda_cv_pred <- predict(lda_cv_fit, newdata = fold_test_df)$class
  
  #SVM
  svm_cv_fit <- svm(Condition ~., data = fold_train_df, kernel = "sigmoid", scale = FALSE)
  svm_cv_pred <- predict(svm_cv_fit, newdata = fold_test_df)
  
  # cross val results 
  crossvalidation[[length(crossvalidation) +1]] <- prediction_metrics(lda_cv_pred, fold_test_df$Condition, "LDA") 
  crossvalidation[[length(crossvalidation) +1]] <- prediction_metrics(svm_cv_pred, fold_test_df$Condition, "SVM Sigmoid")
}

crossval_results_df <- bind_rows(crossvalidation) # performance metrics into one dataframe for plotting
gene_df <- bind_rows(gene_log) # and the selected features/genes into df

ggplot(crossval_results_df, aes(x=Model, y=Accuracy)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "Classifier Type", y = "Accuracy")
  

# assumptions vary between LDA and SVM, svm is presenting as more accuracte (using sigmoid kernel), it is less sensitive to changes in the training set, it will give a more accurte result with the small number of samples compared to LDA
# sigmoid svm rather than linear svm
# svm uses scaling and overfitting is less likely 
```
